---
title: "Implementation of a Neural Network for Iris dataset classification in R"
author: "Introduction to Statistical Learning"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

# Implementation of a Neural Network for Iris dataset classification in R

This notebook implements a simple multilayer neural network for classifying flower species in the Iris dataset.

The code is modularized into functions to facilitate understanding and step-by-step execution, following the same structure as the Python lab.

## Introduction

In this lab we build a simple Neural Network, implementing every function from scratch in R.

The problem to be solved is the identification (*classification*, *prediction*) of the species of an Iris flower using the dimensions of the petals and sepals.

We follow the usual workflow of a machine learning analysis:

1. Load and preprocess the data.
2. Build the functions for the neural network.
3. Train the model.
4. Evaluate the performance on test data.

Many of these functions are available in high-level libraries, but here they are written from scratch for illustration purposes.

## Importing libraries

```{r}
# This notebook only uses base R functions.
# ggplot2 is used only for nicer plots if available.
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
}
```

## Load and preprocess data

### Reading the data

```{r}
# Read data from csv.
# The file must be in the working directory.
iris <- read.csv("IrisData1.csv", stringsAsFactors = FALSE)

selected_rows <- c(1, 2, 51, 52, 101, 102)
iris[selected_rows, ]
```

### Dataset verification

A standard step in the data analysis pipeline is to check the data, in order to detect possible problematic values such as missing data, outliers, or coding mistakes.

```{r}
verify_dataset <- function(data) {
  if (any(is.na(data))) {
    stop("Data contain missing values.")
  }
  cat("Dataset is complete. No missing values. Ok\n")
  cat("Class labels:\n")
  print(table(data$Species))
}

verify_dataset(iris)
```

### One-hot encoding function

The target variable `Species` has three categorical classes:

- `Iris-setosa`
- `Iris-versicolor`
- `Iris-virginica`

Neural networks require numerical outputs. We therefore transform the class labels into a one-hot encoded matrix.

```{r}
to_one_hot <- function(labels) {
  labels <- as.factor(labels)
  classes <- levels(labels)
  y <- matrix(0, nrow = length(labels), ncol = length(classes))
  colnames(y) <- classes
  y[cbind(seq_along(labels), as.integer(labels))] <- 1
  return(y)
}

# Example of one-hot encoding for selected rows
to_one_hot(iris$Species[selected_rows])
```

### Data normalization function

The training of neural networks can be negatively affected by input features with very different scales.

To keep this R lab aligned with the Python lab, we use row-wise L2 normalization: each observation is divided by its Euclidean norm.

```{r}
normalize_rows <- function(X) {
  X <- as.matrix(X)
  norm <- sqrt(rowSums(X^2))
  norm[norm == 0] <- 1
  X / norm
}
```

### Running the preprocessing steps

```{r}
# Select numeric features
selected_features <- c("Sepal_Length", "Sepal_Width", "Petal_Length", "Petal_Width")
X <- as.matrix(iris[, selected_features])
X <- normalize_rows(X)

# One-hot encode the outcome
Y <- to_one_hot(iris$Species)

# Show normalized features and encoded outcome for selected rows
X[selected_rows, ]
Y[selected_rows, ]
```

### Test-training split function

```{r}
split_dataset_test_train <- function(X, Y, train_size = 0.7, seed = 123456) {
  set.seed(seed)
  n <- nrow(X)
  idx <- sample(seq_len(n))
  n_train <- floor(train_size * n)
  train_idx <- idx[seq_len(n_train)]
  test_idx <- idx[(n_train + 1):n]
  list(
    X_train = X[train_idx, , drop = FALSE],
    Y_train = Y[train_idx, , drop = FALSE],
    X_test = X[test_idx, , drop = FALSE],
    Y_test = Y[test_idx, , drop = FALSE]
  )
}
```

### Split the data

```{r}
train_test_data <- split_dataset_test_train(X, Y, train_size = 0.7, seed = 123456)

X_train <- train_test_data$X_train
Y_train <- train_test_data$Y_train
X_test <- train_test_data$X_test
Y_test <- train_test_data$Y_test

X_train[1:6, ]
Y_train[1:6, ]
```

## Training process

The network is trained by a succession of forward and backward steps, continuously adjusting the model parameters to minimize the loss.

- **Forward propagation**: computes activations in each layer.
- **Loss computation**: measures the discrepancy between predicted and true values.
- **Backpropagation**: computes gradients by applying the chain rule.
- **Weight update**: adjusts the parameters using gradient descent.

This process is repeated over multiple epochs until convergence or a stopping criterion is met.

## Activation functions

```{r}
sigmoid <- function(x) {
  1 / (1 + exp(-x))
}

sigmoid_deriv_from_activation <- function(a) {
  # If a = sigmoid(z), then sigmoid'(z) = a * (1 - a)
  a * (1 - a)
}

softmax <- function(x) {
  # Numerically stable softmax, applied row-wise
  x_shift <- x - apply(x, 1, max)
  exp_x <- exp(x_shift)
  exp_x / rowSums(exp_x)
}
```

## Forward propagation

Forward propagation computes the activations in each layer of the network.

For a network with one hidden layer:

$$
Z_h = XW_0 + b_0,
\qquad
A_h = \sigma(Z_h)
$$

$$
Z_o = A_h W_1 + b_1,
\qquad
A_o = \operatorname{softmax}(Z_o)
$$

where $A_h$ and $A_o$ are respectively the hidden-layer and output-layer activations.

```{r}
forward_propagation <- function(X, W0, b0, W1, b1) {
  Z_h <- X %*% W0 + matrix(b0, nrow = nrow(X), ncol = length(b0), byrow = TRUE)
  A_h <- sigmoid(Z_h)
  Z_o <- A_h %*% W1 + matrix(b1, nrow = nrow(A_h), ncol = length(b1), byrow = TRUE)
  A_o <- softmax(Z_o)
  list(A_h = A_h, A_o = A_o)
}
```

## Backpropagation

Backpropagation propagates the error backward to adjust weights using gradient descent.

For the output layer with softmax and cross-entropy:

$$
\delta_o = A_o - Y
$$

For the hidden layer:

$$
\delta_h = (\delta_o W_1^T) \odot \sigma'(Z_h)
$$

Since the function below receives $A_h = \sigma(Z_h)$, we use:

$$
\sigma'(Z_h) = A_h(1-A_h)
$$

```{r}
backpropagation <- function(X, Y, A_h, A_o, W1) {
  delta_o <- A_o - Y
  dcost_dah <- delta_o %*% t(W1)
  delta_h <- dcost_dah * sigmoid_deriv_from_activation(A_h)
  list(delta_o = delta_o, delta_h = delta_h)
}
```

## Updating weights and biases

```{r}
update_weights <- function(X, A_h, delta_o, delta_h, W0, b0, W1, b1, learning_rate) {
  W1 <- W1 - learning_rate * (t(A_h) %*% delta_o)
  b1 <- b1 - learning_rate * colSums(delta_o)
  W0 <- W0 - learning_rate * (t(X) %*% delta_h)
  b0 <- b0 - learning_rate * colSums(delta_h)
  list(W0 = W0, b0 = b0, W1 = W1, b1 = b1)
}
```

## Initialization of weights and biases

Before training the network, weights and biases must be initialized.

Weights are randomly initialized to break symmetry in learning.

```{r}
initialize_network <- function(input_size, hidden_size, output_size, seed = 654321) {
  set.seed(seed)
  W0 <- matrix(runif(input_size * hidden_size, min = -1, max = 1),
               nrow = input_size, ncol = hidden_size)
  W1 <- matrix(runif(hidden_size * output_size, min = -1, max = 1),
               nrow = hidden_size, ncol = output_size)
  b0 <- rnorm(hidden_size)
  b1 <- rnorm(output_size)
  list(W0 = W0, b0 = b0, W1 = W1, b1 = b1)
}
```

## Loss function

```{r}
cross_entropy_loss <- function(Y, A_o, eps = 1e-9) {
  -mean(rowSums(Y * log(A_o + eps)))
}
```

## Neural network training

After setting the weights to their initial values, the neural network is trained by iterating over the data.

At each epoch and batch:

1. Compute predictions by forward propagation.
2. Compute errors by backpropagation.
3. Update weights and biases.

The process stops when convergence or a fixed number of epochs is reached.

```{r}
train_neural_network <- function(X_train, Y_train, epochs, learning_rate, batch_size, params) {
  W0 <- params$W0
  b0 <- params$b0
  W1 <- params$W1
  b1 <- params$b1
  
  num_samples <- nrow(X_train)
  num_batches <- ceiling(num_samples / batch_size)
  loss_history <- c()
  
  for (epoch in seq_len(epochs)) {
    for (batch_idx in seq_len(num_batches)) {
      batch_start <- (batch_idx - 1) * batch_size + 1
      batch_end <- min(batch_idx * batch_size, num_samples)
      
      X_batch <- X_train[batch_start:batch_end, , drop = FALSE]
      Y_batch <- Y_train[batch_start:batch_end, , drop = FALSE]
      
      # Forward propagation
      fwd <- forward_propagation(X_batch, W0, b0, W1, b1)
      
      # Backpropagation
      bp <- backpropagation(X_batch, Y_batch, fwd$A_h, fwd$A_o, W1)
      
      # Weight update
      updated <- update_weights(X_batch, fwd$A_h, bp$delta_o, bp$delta_h,
                                W0, b0, W1, b1, learning_rate)
      W0 <- updated$W0
      b0 <- updated$b0
      W1 <- updated$W1
      b1 <- updated$b1
    }
    
    # Compute loss every 100 epochs, and also at the first epoch
    if (epoch == 1 || epoch %% 100 == 0) {
      fwd_all <- forward_propagation(X_train, W0, b0, W1, b1)
      loss <- cross_entropy_loss(Y_train, fwd_all$A_o)
      loss_history <- c(loss_history, loss)
      cat(sprintf("Epoch %d, Loss: %.4f\n", epoch, loss))
    }
  }
  
  list(params = list(W0 = W0, b0 = b0, W1 = W1, b1 = b1),
       loss_history = loss_history)
}
```

## Training the network in practice

### Initialize the network and hyperparameters

We set:

- input layer: 4 nodes (the four Iris measurements)
- hidden layer: 5 nodes
- output layer: 3 nodes (the three species)

Additionally, we set the training hyperparameters:

- learning rate
- batch size
- number of epochs

```{r}
input_size <- ncol(X_train)
hidden_size <- 5
output_size <- ncol(Y_train)

my_params <- initialize_network(input_size, hidden_size, output_size, seed = 654321)

learning_rate <- 0.01
batch_size <- 10
epochs <- 1000
```

### Train the network

```{r}
trained <- train_neural_network(
  X_train = X_train,
  Y_train = Y_train,
  epochs = epochs,
  learning_rate = learning_rate,
  batch_size = batch_size,
  params = my_params
)

trainedNet <- trained$params
loss_history <- trained$loss_history
```

## Check model performance

```{r}
plot(loss_history, type = "l",
     xlab = "Logged training step",
     ylab = "Loss",
     main = "Training Loss over Time")
```

## Evaluate network on test data

```{r}
evaluate_network <- function(params, X_test) {
  fwd <- forward_propagation(X_test, params$W0, params$b0, params$W1, params$b1)
  # Return class index for each row
  max.col(fwd$A_o, ties.method = "first")
}

y_pred <- evaluate_network(trainedNet, X_test)
y_actual <- max.col(Y_test, ties.method = "first")

cm <- table(
  Actual = colnames(Y_test)[y_actual],
  Predicted = colnames(Y_test)[y_pred]
)

cm
```

### Normalized confusion matrix

```{r}
cm_norm <- prop.table(cm, margin = 1)
round(cm_norm, 3)
```

```{r}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  df_cm <- as.data.frame(cm_norm)
  ggplot(df_cm, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile() +
    geom_text(aes(label = round(Freq, 2)), size = 5) +
    labs(title = "Normalized Confusion Matrix", x = "Predicted", y = "Actual") +
    theme_minimal()
} else {
  image(as.matrix(cm_norm), main = "Normalized Confusion Matrix")
}
```

## Predicting new observations

The following function normalizes a new observation using the same row-wise normalization used during training and returns the predicted species.

```{r}
predict_species <- function(params, measurements) {
  measurements <- matrix(as.numeric(measurements), nrow = 1)
  measurements_norm <- normalize_rows(measurements)
  fwd <- forward_propagation(measurements_norm, params$W0, params$b0, params$W1, params$b1)
  predicted_class <- which.max(fwd$A_o[1, ])
  list(
    probabilities = fwd$A_o,
    predicted_species = colnames(Y)[predicted_class]
  )
}

# Example
predict_species(trainedNet, c(5.1, 3.5, 1.4, 0.2))
```

